Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Including feedback meeting Gerko

\begin{abstract}
(\textbf{Rewrite:}) Iterative imputation by chained equations (MICE) is
a widely used tool to accommodate missing data. While it is widely
accepted that valid inferences can be obtained with this technique,
these inferences all rely on algorithmic convergence. There is no
consensus on how to evaluate the convergence properties of the method.
This paper provides insight into identifying non-convergence of
iterative impuation algorithms.
\end{abstract}

\keywords{MICE; convergence}

@@

Anyone who analyzes person-data may run into a missing data problem.
Missing data is not only ubiquitous but also \textbf{tedious}. If a
dataset contains just one incomplete observation, statistical inferences
are undefined and will not produce any results. To circumvent this, many
statistical packages employ list-wise deletion by default (i.e.,
ignoring incomplete observations). Unfortunately, this \emph{ad hoc}
solution may yield wildly invalid results \citep{buur18}. An alternative
is to `impute' (i.e., fill in) the missing values in the incomplete
observations. Subsequently, statistical inferences can be performed on
the completed dataset. By repeating this process several times, a
distribution of plausible results may be obtained, which reflects the
uncertainty in the data due to missingness. This technique is known as
`multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a
powerful technique to yield unbiased and confidence valid estimates of
the true---but missing---data inference under many circumstances
\citep{buur18}.

\textbf{Move this after terminology??} Figure \ref{fig:diagram} provides
an overview of the steps involved with MI---from incomplete data, to
\(M\) completed datasets, to \(M\) estimated quantities of interest
(\(\hat{Q}\)s), to a single pooled estimate \(\bar{Q}\). Missing data in
\(y\) is `imputed' (i.e., filled in) \(M\) times. The imputed data
(\(y_{imp}\)) is combined with the observed data (\(y_{obs}\)) to create
\(M\) completed datasets. On each completed dataset, the analysis of
scientific interest is performed. The quantity of scientific interest
(e.g., a regression coefficient) is denoted with \(Q\). Since \(Q\) is
estimated on each completed dataset, \(M\) separate \(\hat{Q}\)-values
are obtained. These \(M\) values are combined into a single pooled
estimate \(\bar{Q}\).

@@

Currently, the recommended practice for evaluating convergence is to
visually inspect imputations for signs of non-convergence. This method
is insufficient on two counts: 1) it may be challenging to the untrained
eye, and 2) only severely pathological cases of non-convergence may be
diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative,
diagnostic evaluation of convergence would be preferred.

Monitoring convergence of iterative imputation algorithms diagnostically
is challenging. Iterative imputation algorithms such as MICE are Markov
chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not
from a scalar to a point but from one distribution to another. The
values generated by the algorithm (e.g., imputed values) will vary even
after convergence. Therefore, the aim of convergence diagnostics for
MCMC methods is not to establish the point at which convergence is
reached, but to monitor signs of non-convergence. Several of such
diagnostics exist for MCMC methods, but it is not known whether these
are appropriate for MICE.

In this paper, we investigate how non-convergence in iterative
imputation algorithms may be diagnosed, how well these methods perform,
and at which point convergence may safely be assumed. For reasons of
brevity, we only focus on the MI algorithm implemented in the popular
\texttt{mice} package \citep{mice} in \texttt{R} \citep{R}. The
convergence properties of the MICE algorithm are investigated through
model-based simulation. The results of this simulation study are
guidelines for assessing convergence of MI algorithms, which will aid
applied researchers in drawing valid inference from incomplete datasets.

\textbf{Include sub-questions?:} How can non-convergence be diagnosed?
Are common MCMC non-convergence diagnostics appropriate for MICE? And if
so, which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
has not converged? And what are guidelines for practice?}

\hypertarget{terminologynotation}{%
\subsection{Terminology/Notation}\label{terminologynotation}}

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
(\(j = 1, 2, \dots, p\)) may be either observed or missing. The
collection of observed data values in \(y\) is denoted by \(y_{obs}\);
the missing part of \(y\) is referred to as \(y_{mis}\). For each
datapoint in \(y_{mis}\), we sample \(M \times T\) times plausible
values, where \(M\) is the number of imputations
(\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations
(\(t = 1, 2, \dots, T\)). The collection of samples between the initial
value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be
referred to as an `imputation chain'. \textbf{Add: missingness
percentage = proportion of cases with one or more missing values, Q is
quantity of scientific interest: ``A scientific estimand \(Q\) is a
quantity of scientific interest that we can calculate if we would
observe the entire population'' \citep[par 2.3.1]{buur18}, and
\(\theta\)s are scalar summaries of interest in the iterative algorithm
(e.g., chain means; the average of the imputed values in each imputation
chain). }

@@

Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, mixing and stationarity will be evaluated
separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
\(AC\) and \(\widehat{R}\) will be used \textbf{to diagnose
non-convergence}.

@@

To define \(\widehat{R}\), we follow notation by \citep[p.~5]{veht19}.
Let \(M\) be the total number of chains, \(T\) the number of iterations
per chain, and \(\theta\) the scalar summary of interest (e.g., chain
mean or chain variance). For each chain (\(m = 1, 2, \dots, M\)), we
estimate the variance of \(\theta\), and average these to obtain
within-chain variance \(W\).

@@

\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
the chain means are plotted again, now together---as are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up }in the typical convergence situation\textbf{. The
superior performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function
\texttt{stats::acf()}. When we look at this panel, we conclude something
weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
the typical convergence and the pathological non-convergence, while
there is obvious trending in the latter. Moreover, the best convergence
(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf() \citep{R}, we conclude that this
implementation of \(AC\) is not suitable for iterative imputation
algorithms. There is a correction factor for a mathematical shortcut
that works in the limit. According to \citet{box15}, the function works
when the number of iterations \textgreater{} 50. The default number of
iterations in iterative imputation, however, is often lower. Therefore,
we compute \(AC\) manually, see panel D. The AC values in this plot do
meet the requirements (\ldots) and will therefore be used in the
simulation study.}

@@

\hypertarget{simulation-hypothesis}{%
\subsection{Simulation Hypothesis}\label{simulation-hypothesis}}

@@

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. The threshold of \(\widehat{R} < 1.01\) will then
be too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
values are chosen randomly from the observed data. Therefore, we cannot
be certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\)
\textbf{explain that the randomness induced by the MICE algorithm would
take care of this??}. \textbf{(Add actual hypothesis: we thought that
mice would converge sooner than \(\widehat{R}\) would indicate that it
did)} No hypothesis was formulated about the performance of \(AC\) as
convergence diagnostic. \textbf{Added later: High \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains.
But this doesn't seem the case in practice??}

@@

The two challenges of convergence are stationarity and mixing. Mixing
and stationarity can be inspected visually, by evaluating traceplots of
scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
variances). As \citet{buur18} describes, users can also specify a
model-specific scalar summary (e.g., a regression coefficient). A
user-specified scalar summary, however, is not universal to all complete
data problems. Therefore, as inspired by \citep{mack03}, we will propose
a \(\theta\) that summarizes the multivariate state space of the
algorithm. Namely, the first eigenvalue of the variance-covariance
matrix of the \(M\) completed datasets. \textbf{Explain
model-dependent\ldots{}}

The methods under evaluation are \(\widehat{R}\) and \(AC\). These will
be computed for the three scalar summaries of interest: chain means
(i.e., per imputation, per variable: on \(y_{imp, m, j}\)), chain
variances (i.e., per imputation, per variable: on \(y_{imp, m, j}\)),
and the first eigenvalue of the variance-covariance matrix across
imputations (i.e., per imputation: on \{\(y_{obs}, y_{imp, m}\)\})
\textbf{add user-specified \(\theta\): estimated regression coefficient
per imputation}.

@@

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
thresholds for MCMC algorithms, because \textbf{it's} convergence to a
distribution. In practice, non-convergence is usually diagnosed when
\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
performed to assess whether \(AC\) is significantly different from zero.
Performance measures are the same as above: unbiased, confidence valid
estimates.

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence by: 1) terminating the MICE algorithm at different
imputation chain lengths (\(t = 1, 2,..., 100\)), and 2) varying the
missingness proportions (\(p_{miss} =.05,.25,.50,.75,.95\)). The
assumption underlying the different number of iterations is that the
algorithm generally does reach convergence at \(t=1\), because the
initial values are sampled randomly from the set of observed datapoints.
\textbf{As the number of iterations goes up, the imputation chains will
become independent of the initial values until the point at which an
extra added iteration does not lead to a more converged state.} The
second set of experimental conditions under consideration--the
missingness proportions--are chosen to reflect the difficulty of the
missingness problem. The inherent assumption is that low missingness
proportions will lead to quick algorithmic convergence, since there is a
lot of information in the observed data. Higher missingness proportions
then cause slower convergence. However, if the fraction of missing
information is very high, there is so little information in the data
that the random component in the algorithm will take the overhand and a
stable but very uncertain (high variance) point will be reached.

The data-generating mechanism is a multivariate normal distribution,
representing person data on three independent variables (from an
unspecified social scientific field of study). Let \textbf{what?}

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} In this study, sampling variance is not of interest.
Therefore, a single complete set may serve as comparative truth in all
simulation runs \citep{vink14}. A finite population of \(N=1000\) is
simulated using the \texttt{mvtnorm} package \citep{mvtnorm}.
Subsequently, a fourth variable is constructed to serve as dependent
variable in a multiple linear regression problem. Let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

\textbf{We consider four quantities of scientific interest
\citep[`conceptual estimands';][]{morr19}, namely the descriptive
statistics of all variables (mean and standard deviation), \(\beta_2\),
and the percentage of variance in \(Y\) explained (coefficient of
determination; \(R^2\)). We solve a multiple linear regression problem,
where dependent variable \(Y\) is regressed on independent variables
\(X_1\), \(X_2\) and \(X_3\)}

@@

Missing datapoints in \(y\) are imputed with the \texttt{mice} package
\citep{mice}. All MI procedures are performed with Bayesian linear
regression imputation (\texttt{method\ =\ "norm"}), and five imputation
chains (\texttt{m\ =\ 5}). The number of iterations varies between
simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)). Each
repetition of the simulation thus starts from 1 complete dataset, then
splits into 5 missingness conditions (\(p =.05,.25,.5,.75,.95\)), each
of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\)) for
each iteration condition (\(t = 1, 2,..., 100\)). From these
\(5\times5\times100\) imputations, we extract the \(\theta\)s to apply
\(\widehat{R}\) and \(AC\) on (\textbf{i.e., chain means, chain
variances, \(\beta\)s per imputation, and the first eigenvalue of the
variance-covariance matrix per imputation}). We calculate
\(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19}
recommendations, and \(AC\) as the correlation between the \(t^{th}\)
and the \((t+1)^{th}\) iteration. For each \(\theta\) we estimate the
corresponding scientific estimand \(Q\).

The estimator for each \(Q\) is \(\bar{Q}\)---the pooled aggregate of
the (\(\hat{Q}\))s across imputations. To calculate the \(\bar{Q}\)s, we
combine the observed data (\(y_{obs}\)) and the imputed data for each
imputation (\(y_{imp,m}\)) with the function \texttt{mice::complete()}.
Descriptive statistics are computed as the average across imputations
for each variable (\(\mu_j\) and \(\sigma_j\), where
\(j = Y, X_1, X_2, X_3\)). Estimated regression coefficients are
obtained with the function \texttt{stats::lm()} for each imputation, and
then pooled conform \citet{vink14}. The coefficient of determination is
estimated for each imputation, and pooled using
\texttt{mice::pool.r.squared()}.

>>>>

\begin{abstract}
Iterative imputation is a popular tool to accommodate missing data.
While it is widely accepted that valid inferences can be obtained with
this technique, these inferences all rely on algorithmic convergence.
There is no consensus on how to evaluate the convergence properties of
the method. This paper provides insight into identifying non-convergence
of iterative impuation algorithms.
\end{abstract}

\keywords{iterative imputation; non-convergence; mice}

@@

Anyone who analyzes person-data may run into a missing data problem.
Missing data is not only ubiquitous, but treating it can also be
tedious. If a dataset contains just one incomplete observation,
statistical inferences are undefined and will not produce any results.
To circumvent this, many statistical packages employ list-wise deletion
by default (i.e., ignoring incomplete observations). Unfortunately, this
\emph{ad hoc} solution may yield wildly invalid results \citep{buur18}.
An alternative is to \emph{impute} (i.e., fill in) the missing values in
the incomplete observations. Subsequently, statistical inferences can be
performed on the completed dataset. By repeating this process several
times, a distribution of plausible results may be obtained, which
reflects the uncertainty in the data due to missingness. This technique
is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven
to be a powerful technique to yield unbiased and confidence valid
estimates of the true---but missing---data inference under many
circumstances \citep{buur18}.

Figure \ref{fig:diagram} provides an overview of the steps involved with
MI---from incomplete data, to \(M\) multiply imputed datasets, to \(M\)
estimated quantities of interest \(\hat{Q}\)s, to a single pooled
estimate \(\bar{Q}\). Missing data in dataset \(y\) is imputed \(M\)
times. The imputed data \(y_{imp}\) is combined with the observed data
\(y_{obs}\) to create \(M\) completed datasets. On each completed
dataset, the analysis of scientific interest is performed. The quantity
of scientific interest (e.g., a regression coefficient) is denoted with
\(Q\). Since \(Q\) is estimated on each completed dataset, \(M\)
separate \(\hat{Q}\)-values are obtained. These \(M\) values are
combined into a single pooled estimate \(\bar{Q}\).

@@

Currently, the recommended practice for evaluating convergence in
iterative imputation algorithms is to visually inspect imputations for
signs of non-convergence. \textbf{(Add chain means and variances here
already? Because it's insufficient that these are univariate?)}. This
method is insufficient on two counts: 1) it may be challenging to the
untrained eye, and 2) only severely pathological cases of
non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
Therefore, a quantitative, diagnostic evaluation of convergence would be
preferred. Yet, monitoring convergence of iterative imputation
algorithms diagnostically is challenging. Iterative imputation
algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In
MCMC methods, convergence is not from a scalar to a point but from one
distribution to another. The values generated by the algorithm (e.g.,
imputed values) will vary even after convergence. Therefore, the aim of
convergence diagnostics for MCMC methods is not to establish the point
at which convergence is reached, but to monitor signs of
non-convergence. Several of such diagnostics exist for MCMC methods, but
it is not known whether these are appropriate for iterative imputation
algorithms.

In this paper, we investigate how non-convergence in iterative
imputation algorithms may be diagnosed, how well these methods perform,
and at which point convergence may safely be assumed. For reasons of
brevity, we only focus on the iterative imputation algorithm implemented
in the popular \texttt{mice} package \citep{mice} in \texttt{R}
\citep{R}. The convergence properties of the MICE algorithm are
investigated through model-based simulation. The results of this
simulation study are guidelines for assessing convergence of MI
algorithms, which will aid applied researchers in drawing valid
inference from incomplete datasets.

\textbf{Include sub-questions?:} How can non-convergence be diagnosed?
Are common MCMC non-convergence diagnostics appropriate for MICE? And if
so, which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
has not converged? And what are guidelines for practice? Can the
parameter of interest, estimand \(Q\), be correct when the algorithm is
not (yet) converged, and vice versa?}

\hypertarget{some-notation}{%
\subsection{Some notation}\label{some-notation}}

Let \(y\) denote an \(n \times k\) matrix containing the data values on
\(k\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
(\(j = 1, 2, \dots, k\)) may be either observed or missing. \textbf{The
number of units \(i\) with at least one missing data value can be
divided by the total number of units \(n\) to obtain the `missingness
proportion' \(p_{mis}\).} The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
\(M \times T\) times plausible values, where \(M\) is the number of
imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
will be referred to as an `imputation chain'. \textbf{Add: \(\theta\)s
are scalar summaries of interest in the iterative algorithm (e.g., chain
means; the average of the imputed values in each imputation chain). }

@@

Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, use we \(AC\) and \(\widehat{R}\) to evaluate
stationarity and mixing separately, as recommended by e.g.,
\citet{cowl96}, p.~898.

@@

An updated version of \(\widehat{R}\) has been proposed by
\citet{veht19}, (p.~5). \textbf{This version may be suitable for
iterative imputation.} Let \(M\) be the total number of chains, \(T\)
the number of iterations per chain, and \(\theta\) the scalar summary of
interest (e.g., chain mean or chain variance). For each chain
(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
average these to obtain within-chain variance \(W\).

@@

\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
the chain means are plotted again, now together---as are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up }in the typical convergence situation\textbf{. The
superior performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function
\texttt{stats::acf()}. When we look at this panel, we conclude something
weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
the typical convergence and the pathological non-convergence, while
there is obvious trending in the latter. Moreover, the best convergence
(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf() \citep{R}, we conclude that this
implementation of \(AC\) is not suitable for iterative imputation
algorithms. There is a correction factor for a mathematical shortcut
that works in the limit. According to \citet{box15}, the function works
when the number of iterations \textgreater{} 50. The default number of
iterations in iterative imputation, however, is often much lower.
Therefore, we compute \(AC\) manually, see panel D. The AC values in
this plot do meet the requirements (\ldots) and will therefore be used
in the simulation study.}

\textbf{acf is niet bedoeld om weinig iteraties te\ldots{} niet een fout
in acf, alleen hier niet van toepassing}

@@

\hypertarget{simulation-hypothesis-als-bullets}{%
\subsection{Simulation Hypothesis (als
bullets!!)}\label{simulation-hypothesis-als-bullets}}

@@

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. Lacerda concluded that rhat dropped below 1.2
after at least 50 iterations. Therefore\ldots{} Vehtari's proposed
threshold of \(\widehat{R} < 1.01\) will then be too stringent for
diagnosing convergence. This over-estimation may, however, be diminished
because \(\widehat{R}\) can falsely diagnose convergence if initial
values of the algorithm are not appropriately over-dispersed
\citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are chosen
randomly from the observed data. Therefore, we cannot be certain that
the initial values are over-dispersed. We expect this to have little
effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
actual hypothesis: we thought that mice would converge sooner than
\(\widehat{R}\) would indicate that it did)} We hypothesize that high AC
values are implausible in converged MI algorithms. \textbf{Added later:
High \(AC\)-values are implausible in MI procedures. That is, the
randomness induced by the MI algorithm effectively mitigates the risk of
dependency within chains. But this doesn't seem the case in practice??}

@@

\textbf{Move to section above} Mixing and stationarity have histroically
been inspected visually, by evaluating traceplots of scalar summaries of
interest (\(\theta\)s; e.g., chain means and chain variances). As
\citet{buur18} describes, users can also specify a model-specific scalar
summary (e.g., a regression coefficient). A user-specified scalar
summary, however, is not universal to all complete data problems.
Therefore, as inspired by \citep{mack03}, we propose a \(\theta\) that
summarizes the multivariate state space of the algorithm. Namely, the
first eigenvalue of the variance-covariance matrix of the \(M\)
completed datasets. \textbf{The first eigenvalue has the appealing
property that is not dependent on the substantive model of interest.}

The methods under evaluation are \(\widehat{R}\) and \(AC\). These will
be computed for the \textbf{four} scalar summaries of interest: chain
means (i.e., per imputation, per variable: on \(y_{imp, m, j}\)), chain
variances (i.e., per imputation, per variable: on \(y_{imp, m, j}\)),
and the first eigenvalue of the variance-covariance matrix across
imputations (i.e., per imputation: on \{\(y_{obs}, y_{imp, m}\)\})
\textbf{add user-specified \(\theta\): estimated regression coefficient
per imputation}.

@@

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
thresholds for MCMC algorithms, because of its convergence to a
distribution. In practice, non-convergence is usually diagnosed when
\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
performed to assess whether \(AC\) is significantly different from zero.
Performance measures are the same as above: unbiased, confidence valid
estimates.

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence by: 1) terminating the MICE algorithm at different
imputation chain lengths (\(t = 1, 2,..., 100\)), and 2) varying the
missingness proportions (\(p_{mis} =.05,.25,.50,.75,.95\)). The
assumption underlying the different number of iterations is that the
algorithm generally does reach convergence at \(t=1\), because the
initial values are sampled randomly from the set of observed datapoints.
\textbf{As the number of iterations goes up, the imputation chains will
become independent of the initial values until the point at which an
extra added iteration does not lead to a more converged state.} The
second set of experimental conditions under consideration--the
missingness proportions--are chosen to reflect the difficulty of the
missingness problem. The inherent assumption is that low missingness
proportions will lead to quick algorithmic convergence, since there is a
lot of information in the observed data. Higher missingness proportions
then cause slower convergence. However, if the fraction of missing
information is very high, there is so little information in the data
that the random component in the algorithm will take the overhand and a
stable but very uncertain (high variance) point will be reached.

The data-generating mechanism is a multivariate normal distribution,
representing person data on three predictor variables (from an
unspecified social scientific field of study). Let our predictor space
be defined by the following a multivariate normal distribution

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} In this study, sampling variance is not of interest.
Therefore, a single complete set may serve as comparative truth in all
simulation runs \citep{vink14}. A finite population of \(N=1000\) is
simulated using the \texttt{mvtnorm} package \citep{mvtnorm}.
Subsequently, a fourth variable is constructed to serve as outcome
variable in a multiple linear regression problem. Let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

\textbf{We consider four quantities of scientific interest
\citep[`conceptual estimands';][]{morr19}, namely the descriptive
statistics of all variables (mean and standard deviation), \(\beta_2\),
and the percentage of variance in \(Y\) explained (coefficient of
determination; \(R^2\)). We solve a multiple linear regression problem,
where outcome \(Y\) is regressed on predictors \(X_1\), \(X_2\) and
\(X_3\)}

@@

Missing datapoints in \(y\) are imputed with the \texttt{mice} package
\citep{mice}. All MI procedures are performed with Bayesian linear
regression imputation (\texttt{method\ =\ "norm"}), and five imputation
chains (\texttt{m\ =\ 5}). The number of iterations varies between
simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)). Each
repetition of the simulation thus starts from 1 complete dataset, then
splits into 5 missingness conditions (\(p_{mis} =.05,.25,.5,.75,.95\)),
each of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\))
for each iteration condition (\(t = 1, 2,..., 100\)). From these
\(5\times5\times100\) imputations, we extract the \(\theta\)s to apply
\(\widehat{R}\) and \(AC\) on (\textbf{i.e., chain means, chain
variances, \(\beta\)s per imputation, and the first eigenvalue of the
variance-covariance matrix per imputation}). We calculate
\(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19}
recommendations, and \(AC\) as the correlation between the \(t^{th}\)
and the \((t+1)^{th}\) iteration. For each \(\theta\) we estimate the
corresponding scientific estimand \(Q\).

The estimator for each \(Q\) is \(\bar{Q}\)---the pooled aggregate of
the \(\hat{Q}\)s across imputations. To calculate the \(\bar{Q}\)s, we
combine the observed data \(y_{obs}\) and the imputed data for each
imputation \(y_{imp,m}\) with the function \texttt{mice::complete()}.
Descriptive statistics are computed as the average across imputations
for each variable (\(\mu_j\) and \(\sigma_j\), where
\(j = Y, X_1, X_2, X_3\)). Estimated regression coefficients are
obtained with the function \texttt{stats::lm()} for each imputation, and
then pooled conform \citet{vink14}. The coefficient of determination is
estimated for each imputation, and pooled using
\texttt{mice::pool.r.squared()}.
